SIMD\slash SIMT-vectorized Sparse Chemical Kinetic Jacobian\slash Thermo-chemical Source Term Evaluation

### 2018-03-05 paper.tex
edits & allow per-platform error reporting

Recognizing the need to accelerate Jacobian evaluation and factorization, a number of recent works have been published on analytical chemical kinetic Jacobian evaluation; although as will be discussed at the end of this section, this work offers several key improvements over past efforts.
The \texttt{TChem} software package~\cite{Safta:2011vn} was one of the first packages developed providing this functionality, but has several limitations including: incompatibilities with modern reaction types (i.e., pressure dependent Arrhenius and Chebyshev reactions), and lack of thread-safety to enable parallel execution~\cite{Curtis2017:tchem}. 
Youssefi~\cite{Youssefi:2011tm} explored the importance of analytical Jacobian matrices for time-scale analysis techniques as well as their effect on computational efficiency in zero-dimensional homogeneous reactor simulations.
Bisetti~\cite{Bisetti:2012jw} developed an isothermal, isobaric analytical Jacobian code-generation utility; although incompatible with many combustion codes, a significant increase in Jacobian sparsity was exhibited.
Additionally, in the same work Bisetti provided a novel way to compute dense matrix-vector multiplications resulting from a change of system variables without need for storage of the full dense Jacobian.
Perini et al.~\cite{Perini:2012gy} developed an analytical Jacobian code for constant-volume combustion, with additional options to increase sparsity (at expect of strict correctness), and enable tabulation of temperature dependent properties; a \SI{80}{\percent} speedup over a finite-difference-based Jacobian was reported when used in a multidimensional reactive-flow simulation.
Dijkmans et al.~\cite{Dijkmans:2014bb} developed a GPU-based analytical Jacobian code with optional tabulation of temperature-dependent properties, and found speedups up to \SI{120}{$\times$} for zero-dimensional chemical kinetic integration with large chemical models (\textasciitilde\num{3000} species) with a Tesla C2075 GPU.
Niemeyer et al.~\cite{Niemeyer:2016aa} created and validated the open-source analytical chemical kinetic Jacobian code-generator, \texttt{pyJac}, supporting parallel execution on the CPU and SIMT execution on the GPU; a speedup of \SIrange{3}{7.5}{$\times$} over a finite-difference Jacobian was found on the CPU.
Gao et al.~\cite{GAO2015287} derived a sparse analytical Jacobian, but it was not validated outside the context of use with an implicit-integration technique; in addition the Jacobian was based on an over-constrained system, and a differential algebraic equation (DAE) solver should be used for strict correctness\slash conservation.

@@

\begin{table}[htbp]
\sisetup{retain-zero-exponent=true}
\centering
\begin{tabular}{@{}l S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1]@{}}
\toprule
\multicolumn{1}{l}{Model} & \multicolumn{1}{l}{\ce{H2}\slash~\ce{CO}\cite{Burke:2011fh}} & \multicolumn{1}{l}{GRI-Mech.~3.0\cite{smith_gri-mech_30}} & \multicolumn{1}{l}{USC-Mech II\cite{Wang:2007}} & \multicolumn{1}{l}{\ce{iC5H11OH}\cite{Sarathy:2013jr}} \\
\midrule
\multicolumn{1}{c}{$E_{R^{\prime}}$}                    & 1.56e-8 & 2.95e-8 & 9.42e-8 & 4.86e-4 \\
\multicolumn{1}{c}{$E_{R^{\prime\prime}}$}              & 6.92e-8 & 6.53e-8 & 1.20e-7 & 5.07e-4 \\
\multicolumn{1}{c}{$E_{R}$}                             & 1.49e1  & 1.11e0  & 2.80e0  & 4.82e-1 \\
\multicolumn{1}{c}{$E_{\varepsilon}$}                   & 1.48e1  & 1.13e0  & 2.93e0  & 5.03e-1 \\
\multicolumn{1}{c}{$E_{\frac{\text{d} n}{\text{d} t}}$} & 2.53e1  & 2.60e0  & 7.62e0  & 1.58e1 \\
\multicolumn{1}{c}{$E_{\frac{\text{d} T}{\text{d} t}}$} & 3.94e5  & 3.35e8  & 3.95e6  & 7.11e7 \\
\multicolumn{1}{c}{$E_{\frac{\text{d} S}{\text{d} t}}$} & 3.52e12 & 3.46e12 & 3.44e12 & 3.38e12 \\
\bottomrule
\end{tabular}
\caption{Summary of rate of progress, species and temperature rate correctness.
Error statistics are based on the infinity-norm of the relative error detailed in Eq.~\eqref{e:rel_err} for each quantity.
The ``S'' in $E_{\frac{\text{d} S}{\text{d} t}}$ refers to the thermodynamic state parameter, either $V$ or $P$ for \conp/ and \conv/ respectively.
}
\label{T:source_error}
\end{table}

@@

As in our previous work~\cite{Niemeyer:2016aa}, the correctness of the Jacobian codes generated by \texttt{pyJac} was determined by comparison to the Jacobian obtained by automatic differentiation of the source rate subroutines generated by \texttt{pyJac} via the \texttt{Adept} software library~\cite{adept-v11,hogan2014fast}.
The reasons for this choice are explained fully in~\cite{Niemeyer:2016aa}, but broadly speaking automatic differentiation provides relatively fast, highly accurate Jacobian matrix evaluation with minimal additional programming little effort.
The discrepancy between the analytical and automatic differentiated Jacobians for thermo-chemical state $k$, denoted by $\mathcal{J}_k$ and $\hat{\mathcal{J}}_k$ respectively, is determined by the relative error Frobenius norm over all Jacobian indicies $i, j$:
\begin{equation}
 \label{e:jac_error_base}
 E_{\text{rel}, k} = \frac{\left\lVert \hat{\mathcal{J}}_{ijk} - \mathcal{J}_{ijk} \right\rVert_{i,j,F}}{\left\lVert \hat{\mathcal{J}}_{ijk} \right\rVert_{i,j,F}}
\end{equation}

@@

\begin{table}[htbp]
\centering
\begin{tabular}{@{}l S S[table-format=1.3e1] S[table-format=1.3e1] S[table-format=1.3e1]@{}}
\toprule
Model                 & \multicolumn{1}{c}{Sample size}   & \multicolumn{1}{c}{$\bar{\mathcal{T}}$}  & \multicolumn{1}{c}{$E_{10^{20}}$}   & \multicolumn{1}{c}{$E_{10^{15}}$} \\
\midrule
\ce{H2}\slash \ce{CO} & \num{900900}                      & \num{6.431e+18}      & \num{1.742e00}  & \num{4.508e-5} \\
GRI-Mech 3.0          & \num{450900}                      & \num{7.783e+19}      & \num{3.842e-7}  & \num{3.687e-7} \\
USC-Mech II           & \num{91800}                       & \num{2.830e+21}      & \num{1.119e-2}  & \num{1.983e-7} \\
\ce{iC5H11OH}         & \num{450900}                      & \num{2.733e+26}      & \num{1.363e-3}  & \num{2.764e-5} \\
\bottomrule
\end{tabular}
\caption{Summary of Jacobian matrix validation results.
Error statistics are the largest filtered relative error $E_\mathcal{C}$ for over all samples.
It is noted that the threshold described in~\cref{e:thresh} varies slightly between the \conp/ and \conv/ cases; the reported $\bar{\mathcal{T}}$ is the average of the two, however the appropriate value was used during calculations of the error statistics.
}
\label{T:error}
\end{table}

>>>>

Recognizing the need to accelerate Jacobian evaluation and factorization, a number of recent works have been published on analytical chemical kinetic Jacobian evaluation; although as will be discussed at the end of this section, this work offers several key improvements over past efforts.
The \texttt{TChem} software package~\cite{Safta:2011vn} was one of the first packages developed providing this functionality, but has several limitations including: incompatibilities with modern reaction types---i.e., pressure dependent Arrhenius (or P-Log) and Chebyshev reactions---and lack of thread-safety to enable parallel execution~\cite{Curtis2017:tchem}. 
Youssefi~\cite{Youssefi:2011tm} explored the importance of analytical Jacobian matrices for time-scale analysis techniques as well as their effect on computational efficiency in zero-dimensional homogeneous reactor simulations.
Bisetti~\cite{Bisetti:2012jw} developed an isothermal, isobaric analytical Jacobian code-generation utility; although incompatible with many combustion codes, a significant increase in Jacobian sparsity was exhibited.
Additionally, in the same work Bisetti provided a novel way to compute dense matrix-vector multiplications resulting from a change of system variables without need for storage of the full dense Jacobian.
Perini et al.~\cite{Perini:2012gy} developed an analytical Jacobian code for constant-volume combustion, with additional options to increase sparsity (at expect of strict correctness), and enable tabulation of temperature dependent properties; a \SI{80}{\percent} speedup over a finite-difference-based Jacobian was reported when used in a multidimensional reactive-flow simulation.
Dijkmans et al.~\cite{Dijkmans:2014bb} developed a GPU-based analytical Jacobian code with optional tabulation of temperature-dependent properties, and found speedups up to \SI{120}{$\times$} for zero-dimensional chemical kinetic integration with large chemical models (\textasciitilde\num{3000} species) with a Tesla C2075 GPU.
Niemeyer et al.~\cite{Niemeyer:2016aa} created and validated the open-source analytical chemical kinetic Jacobian code-generator, \texttt{pyJac}, supporting parallel execution on the CPU and SIMT execution on the GPU; a speedup of \SIrange{3}{7.5}{$\times$} over a finite-difference Jacobian was found on the CPU.
Gao et al.~\cite{GAO2015287} derived a sparse analytical Jacobian, but it was not validated outside the context of use with an implicit-integration technique; in addition the Jacobian was based on an over-constrained system, and a differential algebraic equation (DAE) solver should be used for strict correctness\slash conservation.

@@

\begin{table}[htbp]
\sisetup{retain-zero-exponent=true}
\centering
\begin{tabular}{@{}l S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1]@{}}
\toprule
\multicolumn{1}{l}{Model} & \multicolumn{1}{l}{\ce{H2}\slash~\ce{CO}\cite{Burke:2011fh}} & \multicolumn{1}{l}{GRI-Mech.~3.0\cite{smith_gri-mech_30}} & \multicolumn{1}{l}{USC-Mech II\cite{Wang:2007}} & \multicolumn{1}{l}{\ce{iC5H11OH}\cite{Sarathy:2013jr}} \\
\midrule
\multicolumn{1}{c}{$E_{R^{\prime}}$}                    & 1.56e-8 & 2.95e-8 & 9.42e-8 & 4.86e-4 \\
\multicolumn{1}{c}{$E_{R^{\prime\prime}}$}              & 6.92e-8 & 6.53e-8 & 1.20e-7 & 5.07e-4 \\
\multicolumn{1}{c}{$E_{R}$}                             & 1.49e1  & 1.11e0  & 2.80e0  & 4.82e-1 \\
\multicolumn{1}{c}{$E_{\varepsilon}$}                   & 1.48e1  & 1.13e0  & 2.93e0  & 5.03e-1 \\
\multicolumn{1}{c}{$E_{\frac{\text{d} n}{\text{d} t}}$} & 2.53e1  & 2.60e0  & 7.62e0  & 1.58e1 \\
\multicolumn{1}{c}{$E_{\frac{\text{d} T}{\text{d} t}}$} & 3.94e5  & 3.35e8  & 3.95e6  & 7.11e7 \\
\multicolumn{1}{c}{$E_{\frac{\text{d} S}{\text{d} t}}$} & 3.52e12 & 3.46e12 & 3.44e12 & 3.38e12 \\
\bottomrule
\end{tabular}
\caption{Summary of rate of progress, species, temperature and thermodynamic state parameter rate correctness.
Error statistics are based on the infinity-norm of the relative error detailed in Eq.~\eqref{e:rel_err} for each quantity.
The ``S'' in $E_{\frac{\text{d} S}{\text{d} t}}$ refers to the thermodynamic state parameter, either $V$ or $P$ for \conp/ and \conv/ respectively.
}
\label{T:source_error}
\end{table}

@@

As in our previous work~\cite{Niemeyer:2016aa}, the correctness of the Jacobian codes generated by \texttt{pyJac} was determined by comparison to the Jacobian obtained by automatic differentiation of the source rate subroutines generated by \texttt{pyJac} via the \texttt{Adept} software library~\cite{adept-v11,hogan2014fast}.
The reasons for this choice are explained fully in~\cite{Niemeyer:2016aa}, but broadly speaking automatic differentiation provides relatively fast, highly accurate Jacobian matrix evaluation with minimal additional programming effort.
The discrepancy between the analytical and automatic differentiated Jacobians for thermo-chemical state $k$, denoted by $\mathcal{J}_k$ and $\hat{\mathcal{J}}_k$ respectively, is determined by the relative error Frobenius norm over all Jacobian indicies $i, j$:
\begin{equation}
 \label{e:jac_error_base}
 E_{\text{rel}, k} = \frac{\left\lVert \hat{\mathcal{J}}_{ijk} - \mathcal{J}_{ijk} \right\rVert_{i,j,F}}{\left\lVert \hat{\mathcal{J}}_{ijk} \right\rVert_{i,j,F}}
\end{equation}

@@

\begin{table}[htbp]
\sisetup{retain-zero-exponent=true}
\centering
\begin{tabular}{@{}l S S[table-format=1.3e1] S[table-format=1.3e1] S[table-format=1.3e1]@{}}
\toprule
Model                 & \multicolumn{1}{c}{Sample size}   & \multicolumn{1}{c}{$\bar{\mathcal{T}}$}  & \multicolumn{1}{c}{$E_{10^{20}}$}   & \multicolumn{1}{c}{$E_{10^{15}}$} \\
\midrule
\ce{H2}\slash \ce{CO} & \num{900900}                      & \num{6.431e+18}      & \num{1.742e00}  & \num{4.508e-5} \\
GRI-Mech 3.0          & \num{450900}                      & \num{7.783e+19}      & \num{3.842e-7}  & \num{3.687e-7} \\
USC-Mech II           & \num{91800}                       & \num{2.830e+21}      & \num{1.119e-2}  & \num{1.983e-7} \\
\ce{iC5H11OH}         & \num{450900}                      & \num{2.733e+26}      & \num{1.363e-3}  & \num{2.764e-5} \\
\bottomrule
\end{tabular}
\caption{Summary of Jacobian matrix validation results.
Error statistics are the largest filtered relative error $E_\mathcal{C}$ for over all samples.
It is noted that the threshold described in~\cref{e:thresh} varies slightly between the \conp/ and \conv/ cases; the reported $\bar{\mathcal{T}}$ is the average of the two, however the appropriate value was used during calculations of the error statistics.
}
\label{T:error}
\end{table}
